---
title: 'Mesenchymal Stem Cell CLF Single Cell Analysis'
author: "Erda Alexandria Qorri"
date: "2024-09-20"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

### 1. Data Import

```{r load_libraries, echo=FALSE, message=FALSE}
library(SingleCellExperiment)
library(Seurat)
library(SeuratObject)
library(tidyverse) 
library(Matrix)
library(scales)
library(cowplot) 
library(RCurl) 
library(magrittr) 
library(ggplot2)
library(dplyr) 
library(patchwork) 
library(ggpubr)
```

```{r load_filtered_mtx, echo=TRUE, message=FALSE}

# Takes as input the filtered matrices
filtered_mtx_dir <- "/home/alexia/Desktop/Transcriptomics/group/Not_completed/emese_dog_single_cell/genome_99/"
# filtered_mtx_dir <- "/path/to/filtered_mtx/"

filenames <- c("p2_99_filtered_mtx", "p6_99_filtered_mtx")

seurat_objects <- list()

for (file in filenames) {
  # Read 10X data
  seurat_data <- Read10X(data.dir = file.path(filtered_mtx_dir, file))
  
  # Create a Seurat object
  seurat_object <- CreateSeuratObject(counts = seurat_data,
                                      min.features = 200,
                                      min.cells = 3,
                                      project = file)
  
  # Store the Seurat object in the list
  seurat_objects[[file]] <- seurat_object
}

# Merge Seurat objects
seurat_obj_merged_99 <- merge(x = seurat_objects[["p2_99_filtered_mtx"]],
                              y = seurat_objects[["p6_99_filtered_mtx"]],
                              add.cell.ids = c("P2", "P6"))

# Extract metadata for later use
metadata_99 <- seurat_obj_merged_99@meta.data

# Add Sample ID to each cell
metadata_99 <- metadata_99 %>% 
  mutate(Cells = rownames(metadata_99),
         Sample = case_when(
           str_detect(Cells, "^P2") ~ "P2",
           str_detect(Cells, "^P6") ~ "P6",
 TRUE ~ NA_character_))

metadata_99 <- as.data.frame(metadata_99)

# Update the Seurat object with the modified metadata
seurat_obj_merged_99@meta.data <- metadata_99
```

## Part 1: Exploring Sample Quality

1.1 Number of captured cells

```{r qc, echo=TRUE, message=FALSE}

cell_number_plot <- metadata_99 %>%
   ggplot(aes(x = Sample, fill = Sample)) +
   geom_bar() +
   theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
   ggtitle("Cell Number") +
   scale_fill_manual(values = c("darkgreen", "#ff7f0e"))

print(cell_number_plot)
```

1.2 Number of features (genes) vs Number of UMIs

```{r feature_scatter, echo=TRUE, message=FALSE}

feature_scatter_plot <- FeatureScatter(seurat_obj_merged_99, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

print(feature_scatter_plot)
```

1.3 Exploring Mitochondrial contamination

**Note**: For this project we utilized **Ensembl Release 99** of *Canis lupus familiaris*. Later releases were not used because they lack annotations for the mitochondrial genome. Mitochondrial genes were identified using the regular expression pattern "\^MT". Among the 13 mitochondrial genes in the dog genome, three genes- COX3, NDL4, and ND5—were retrieved separately and included in the total count of mitochondrial genes.

```{r mt_contamination, echo=TRUE, message=FALSE}

library(AnnotationHub)
library(AnnotationDbi)
library(ensembldb)

# Initiatlize AnnotationHub
AH = AnnotationHub()

# Query ENSEMBL
ensembl = query(AH, "ENSEMBL")
grep("Canis", (unique(ensembl$species)), value= TRUE)

# Specify release 99
ensembl_canis <- query(AH, c("EnsDb", "Canis lupus familiaris", 99))

if (length(ensembl_canis) == 0) {
   stop("Unable to retrieve the Ensembl release. Check AnnotationHub command!")
}

edb_clf <- ensembl_canis[["AH78741"]]

annotations_clf <- genes(edb_clf,
                         return.type = "data.frame")

# Extract mitochondrial genes
mito_genes <- annotations_clf %>%
  dplyr::filter(seq_name == "MT" & gene_biotype == "protein_coding")

# Extract genes without the "MT" pattern
gene_counts <- GetAssayData(object = seurat_obj_merged_99, slot = "counts")

mt_spec_genes <- c("COX3" = "\\bCOX3\\b", 
                   "ND4L" = "\\bND4L\\b", 
                   "ND5" = "\\bND5\\b")

for (gene in names(mt_spec_genes)) {
  seurat_obj_merged_99[[paste0("pct.mt", tolower(gene))]] <-
    PercentageFeatureSet(seurat_obj_merged_99,
                         pattern = mt_spec_genes[[gene]],
                         assay = "RNA")
}

# Calculate the total mitochondrial percentage of the genes captured by the MT pattern
seurat_obj_merged_99[["pct.mt"]] <- PercentageFeatureSet(seurat_obj_merged_99, 
                                                         pattern = "^MT-",
                                                         assay = "RNA")

# Update the metadata of the seurat object and calculate the total mitochondrial percentage
metadata_99 <- seurat_obj_merged_99@meta.data

metadata_99 <- metadata_99 %>%
   mutate(total_mt = rowSums(across(starts_with("pct.mt"))))

# Assign updated metadata back to the Seurat object
seurat_obj_merged_99@meta.data <- metadata_99
```

1.4 Quality Analysis of Features, UMI Counts, and Mitochondrial Distribution

```{r vln_plot_featurs_umi_mt, echo=TRUE, message=FALSE}
vln_features_mt_plot <- VlnPlot(seurat_obj_merged_99, features = c("nFeature_RNA", "nCount_RNA", "total_mt"), ncol = 3)

print(vln_features_mt_plot)
```

1.5 Joint Quality Control: Correlation Between Features and UMI Counts with Mitochondrial Read Content Distribution

```{r joint_qc_features_umi, echo=FALSE, message=FALSE}

 joint_feature_qc <- metadata_99 %>% 
  ggplot(aes(x=nCount_RNA, y=nFeature_RNA, color=total_mt)) + 
  geom_point() + 
  scale_colour_gradient(low = "blue", high = "red") +
  stat_smooth(method=lm) +
  scale_x_log10() + 
  scale_y_log10() + 
  theme_classic() +
  geom_vline(xintercept = 650) +
  geom_hline(yintercept = 550) +
  facet_wrap(~Sample)

print(joint_feature_qc)
```

## Part 2: Filtering

```{r filtering, echo=TRUE, message=FALSE}

filtered_seurat_obj_merged_99 <- subset(x = seurat_obj_merged_99, 
                                      subset= (nCount_RNA >= 650) &
                                        (nCount_RNA < 37000) &
                                        (nFeature_RNA >= 550) &
                                        (nFeature_RNA < 6500) &
                                        total_mt < 10)
```

## Post-filtering Quality Control

2.1 Joint QC

```{r, post_filtering_joint_qc, echo=TRUE, message=FALSE}

 filtered_joint_qc <- filtered_seurat_obj_merged_99@meta.data %>% 
  ggplot(aes(x=nCount_RNA, y=nFeature_RNA, color=total_mt)) + 
  geom_point() + 
  scale_colour_gradient(low = "blue", high = "red") +
  stat_smooth(method=lm) +
  scale_x_log10() + 
  scale_y_log10() + 
  theme_classic() +
  geom_vline(xintercept = 650) +
  geom_hline(yintercept = 550) +
  facet_wrap(~Sample)

print(filtered_joint_qc)
```

## Gene-level filtering

```{r, gene_level_filtering, echo=TRUE, message=FALSE}

counts <- GetAssayData(object = filtered_seurat_obj_merged_99, slot = "counts")

filtered_counts <- counts[rowSums(counts > 0) >= 10, ]

filtered_seurat_99_adjusted_gf <- CreateSeuratObject(filtered_counts, meta.data = filtered_seurat_obj_merged_99@meta.data)
# 
# # Visually check the filtering
# 
# filtered_seurat_99_adjusted_gf_plot <- filtered_seurat_99_adjusted_gf@meta.data %>% 
#   ggplot(aes(x=nCount_RNA, y=nFeature_RNA, color=total_mt)) + 
#   geom_point() + 
#   scale_colour_gradient(low = "blue", high = "red") +
#   stat_smooth(method=lm) +
#   scale_x_log10() + 
#   scale_y_log10() + 
#   theme_classic() +
#   geom_vline(xintercept = 650) +
#   geom_hline(yintercept = 550) +
#   facet_wrap(~Sample)

# print(filtered_seurat_99_adjusted_gf_plot)
```

# Part 3: Normalization and Clustering

3.1 Normalization

```{r, normalization, echo=TRUE, message=TRUE}
# Normalize using the default seurat command Normalize
seurat_filtered_norm <- NormalizeData(filtered_seurat_99_adjusted_gf)

# Identify variable features, the number of features we limit at 2000
seurat_filtered_norm <- FindVariableFeatures(seurat_filtered_norm, selection.method = "vst", nfeatures = 2000)

```

3.2 Scaling

```{r, scaling, echo=TRUE, message=TRUE}
# Scale the data to standardize the expression values for each gene across all cells. The default scaling method centers and scales the data to have a mean of zero and a standard deviation of one
seurat_filtered_norm_scaled <- ScaleData(seurat_filtered_norm, features = VariableFeatures(object = seurat_filtered_norm))

```

3.3 Principle Component Analysis

```{r, pca_analysis, echo=TRUE, message=TRUE}
# PCA on the normalized and scaled seurat object
seurat_filtered_norm_scaled <- RunPCA(
   seurat_filtered_norm_scaled,
   features = VariableFeatures(object = seurat_filtered_norm_scaled),
   verbose = FALSE
)

# Visualize the first 2 principle components
pca_plot <- DimPlot(seurat_filtered_norm_scaled, reduction = "pca")

print(pca_plot)

# Elbow plot to aid in the selection of the principal components for the UMAP clustering
elbow_plot <- ElbowPlot(seurat_filtered_norm_scaled)

print(elbow_plot)
```

3.4 UMAP Clustering

Upon visual inspection of the elbow plot, we deemed appropriate to select the first 10 principle components as that it is when the variance reclines.

```{r, umap_clustering, echo=TRUE, message=TRUE}
# First we apply FindNeighbors which identifies the nearest neighbors for each cell based on their gene expression profiles and outputs then as a graph where the cells (nodes) are connected to their nearest neighbors (edges)
seurat_filtered_norm_scaled <- FindNeighbors(seurat_filtered_norm_scaled, dims = 1:10)

# FindClusters was used to group cells with similar expression patterns to identify distinct cell populations or states within the dataset
seurat_filtered_norm_scaled <- FindClusters(seurat_filtered_norm_scaled, resolution = 0.2)

# Apply UMAP clustering to the identified populations
seurat_filtered_norm_scaled <- RunUMAP(seurat_filtered_norm_scaled, dims = 1:10)

# Visualize the UMAP Clustering
# Separated by sample of origin
umap_filtered_plot <- DimPlot(seurat_filtered_norm_scaled, reduction = "umap", group.by = "Sample",
                       pt.size = 0.5,
                       cols = c("#0d1b53", "#fe828c"))

print(umap_filtered_plot)


# All the identified cell populations
umap_populations_plot <- DimPlot(seurat_filtered_norm_scaled, reduction = "umap", label = TRUE,
                label.size = 7,
                pt.size = 0.5,
                label.color = "black",
                cols = c("#ad497b", "#cb98dc", "#a6cd8e", "#3496BB", "#92e3de", "#5c7650", "#e94560", "#FFFF00FD"))

print(umap_populations_plot)
```

# Part 4: Find the correct resolution

To determine the optimal resolution for our dataset, we tested the following resolution values: *0.2, 0.6, 0.8, 1.0, 1.4*.

```{r, determine_resolution, echo=TRUE, message=FALSE}
seurat_filtered_norm_scaled_res_screen <- seurat_filtered_norm_scaled

seurat_filtered_norm_scaled_res_screen <- FindClusters(seurat_filtered_norm_scaled_res_screen, resolution = c(0.2, 0.6, 0.8, 1.0, 1.4))

# Resolution: 0.2 
Idents(object = seurat_filtered_norm_scaled_res_screen) <- "RNA_snn_res.0.2"

resolution_0.2 <- DimPlot(seurat_filtered_norm_scaled_res_screen,
                          reduction = "umap",
                          label.box = TRUE,
                          label = TRUE,
                          label.size = 4,
                          cols = c("#ad497b", "#cb98dc", "#a6cd8e", "#3496BB", "#92e3de", "#5c7650", "#e94560", "#FFFF00FD"))

# print(resolution_0.2)

# Resolution 0.6
Idents(object = seurat_filtered_norm_scaled_res_screen) <- "RNA_snn_res.0.6"

resolution_0.6 <- DimPlot(seurat_filtered_norm_scaled_res_screen,
        reduction = "umap",
        label = TRUE,
        label.size = 6)

# print(resolution_0.6)

# Resolution 0.8
Idents(object = seurat_filtered_norm_scaled_res_screen) <- "RNA_snn_res.0.8"

resolution_0.8 <- DimPlot(seurat_filtered_norm_scaled_res_screen,
                          reduction = "umap",
                          label = TRUE,
                          label.size = 6)

# print(resolution_0.8)

# Resolution 1.0
Idents(object = seurat_filtered_norm_scaled_res_screen) <- "RNA_snn_res.1"

resolution_1.0 <- DimPlot(seurat_filtered_norm_scaled_res_screen,
                          reduction = "umap",
                          label = TRUE,
                          label.size = 6)

# print(resolution_1.0)

# Resolution 1.4
Idents(object = seurat_filtered_norm_scaled_res_screen) <- "RNA_snn_res.1.4"

resolution_1.4 <- DimPlot(seurat_filtered_norm_scaled_res_screen,
                          reduction = "umap",
                          label = TRUE,
                          label.size = 6)

# print(resolution_1.4)

```

# Part 5: Intercluster Comparison, identifying markers

```{r, marker_identification, echo=TRUE, message=TRUE}

# FindMarkers: intercluster comparison 1,2,5,6 vs 0,3,4,7

cluster1256v0347markers <- FindMarkers(filtered_seurat_99_adjusted_gf,

                                       ident.1 = c(1,2,5,6),

                                       ident.2 = c(0,3,4,7),

                                       only.pos = FALSE,

                                       logfc.threshold = 0.25,

                                       min.pct = 0.25)

# FindAllMarkers

combined_markers <- FindAllMarkers(object = filtered_seurat_99_adjusted_gf,

                                   logfc.threshold = 0.25,

                                   only.pos = TRUE)
```
